---
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Chapter 6: Analysis of Longitudinal Data

## Data Wrangling
The R code for the Data is available in the "./data/meet_and_repeat.R"

## Data Analysis
Load libraries needed 
```{r}
lapply(c("dplyr", "ggplot2", "tidyr", "lme4", "tidyverse"), library, character.only = TRUE)
```
```{r}
rats=read_csv("./data/rats_long.csv")
bprs=read_csv("./data/bprs_long.csv")
```
### RATS
We start with the longitudinal analysis of the RATS. This dataset is  from a study on understanding the effects of different diets on rat weights over a period of a few weeks. There are 176 datapoints and 5 variables in the long format of the dataset.
- ID: factor variable and ID of the rat 1-16
- Group: Group of the rat belonging to which diet
- WD: the dat the measurement was taken
- Weight: the weight of the rat in grams
- time
```{r}
colnames(rats)
head(rats)
```
```{r}
summary(rats)
```
```{r}
summary(rats)
rats$Group=factor(rats$Group)
rats$ID=factor(rats$ID)
summary(rats)
```
Rats in Group 1 and group 3 have low within group variablity compared to group 2 that are more wildy dispersed.
Group 1 rats have lower weight than other rats in other groups and group 3 has higher weight. There is also one odd rat in group 2 seems to be an outlier. The overall trend within each group is upwards- as time goes on, rats gain more weigth regardless of which group they belong to. 
```{r}
ggplot(rats, aes(x = Time, y = Weight, linetype = ID, color = Group)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times = 4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(min(rats$Weight), max(rats$Weight)))
```
We will then standardize the variables so that they are easier to compare and do the same plot
```{r}
rats=rats %>%
  group_by(Group) %>%
  mutate(stdWeight = (Weight-mean(Weight))/sd(Weight)) %>%
  ungroup()
```

Now there appears to be an outlier in each group but we still see the same upwards trend.
```{r}
ggplot(rats, aes(x = Time, y = stdWeight, linetype = ID, color = Group)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times = 4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(min(rats$stdWeight), max(rats$stdWeight)))
```
Then we make a summary graph with standard deviation and mean

```{r}
rats_summary =rats %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), sterror = sd(Weight)  ) %>%
  ungroup()
```

```{r}
ggplot(rats_summary, aes(x = Time, y = mean, linetype = Group, shape = Group, color = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1, 2, 3)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(1, 2, 3)) +
  geom_errorbar(aes(ymin = mean - sterror, ymax = mean + sterror, linetype = "1"), width = 0.3) +
  theme(legend.position = c(0.8, 0.4)) +
  scale_y_continuous(name = "mean +- sterr")
```
```{r}
rats1 = rats %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

ggplot(rats1, aes(x = Group, y = mean, fill = Group)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape = 21, size = 3, position = position_dodge(width = 0.75), color = "black") +
  scale_y_continuous(name = "mean") +
  theme_minimal() 
```
```{r}
rats_remove_outliers=filter(rats1, (Group==1 & mean>250) | (Group==2 & mean < 590) | (Group==3 & mean>500))
```

We have removed outliers for each group- and we see that things have changed in the boxplots. Group 2 lost skewbness and so does group 3. Group 1 gained some skewness towards high values 
```{r}
ggplot(rats_remove_outliers, aes(x = Group, y = mean, fill = Group)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape = 21, size = 3, position = position_dodge(width = 0.75), color = "black") +
  scale_y_continuous(name = "mean remove outliers") +
  theme_minimal() 
```

### BPRS
This dataset was taken from 40 males undergoing treatment that assesses hostility, hallucinations and gradiosity which is rated on a brief psychiatric rating scale--BPRS from 1(not severe) to 7(very severe)
360 datapoints and 5 variables:
- subject: 1-20 
- treatment: 1-2 
- weeks: which week the measurment took place
- bprs: score numeric
- week: integer 0-8
```{r}
colnames(bprs)
head(bprs)
```

```{r}
dim(bprs)
```
```{r}
bprs$treatment=as.factor(bprs$treatment)
ggplot(bprs, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(col = treatment))+
  scale_y_continuous(name = "BPRS")+
  theme(legend.position = "top")

```
```{r}
bprs$subject=as.factor(bprs$subject)
ggplot(bprs, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(col = subject))+
  scale_y_continuous(name = "BPRS")+
  theme(legend.position = "top")
```
We can kinda identify a downwards trend happening. Now we do regression analysis. week appears to be highly significant w p value 2e-16. The estimate is -2.27 which indicates that there is a decrease in bprs as week increases which is what we saw before in the plots. Treatment is not significant-- so no support that the 2 treatments influence the BPRS rating. The r2 is also low so these are not good variables. We can remove it and redo the analysis.

```{r}
bprs_regression=lm(bprs~ week + treatment, data=bprs)

summary(bprs_regression)
```
Theres only a little improvement in r2 so week is good but im not impressed by the model.
```{r}
bprs_regression_rm=lm(bprs~ week , data=bprs)

summary(bprs_regression_rm)
```
Lets start with the random intercept model. treatment is still not significant
```{r}
bprs_random=lmer(bprs ~ week + treatment + (1 | subject), data = bprs, REML = FALSE)

summary(bprs_random)
```
Random intercept and random slope model
```{r}
bprs_random_both <- lmer(bprs ~ week + treatment +(week | subject), data = bprs, REML = FALSE)

summary(bprs_random_both)
```
```{r}
anova(bprs_random_both, bprs_random)
```
we see a small p value so bprs_random_both is a better fit 

Now we can look at model with interactions
```{r}
bprs_interaction= lmer(bprs ~ week + treatment + week*treatment + (week | subject), data = bprs, REML = FALSE)
summary(bprs_interaction)

```
now we do anova
```{r}
anova(bprs_interaction,bprs_random_both)
```
becasue of the large p value then no strong implication that an interaction model is better-- so the previous is better

plot the fitted vs the best fit

```{r}

bprs_new =bprs %>% mutate(fitted_v = fitted(bprs_random_both))

ggplot(bprs_new, aes(x = week, y = fitted_v, col = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "week", breaks = seq(0, 8, 2)) +
  scale_y_continuous(name = "Fitted weight (grams)") +
  theme(legend.position = "right")
```

